---
title: "Making cheese"
output: 
    rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Tutorial on cheese}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

Being able to structure code that can get to presentation-ready results in an automated way, and is robust to updates in the data, changes in methodological specfications, etc. can not only be a huge time saver, but also help mitigate errors from tedious tasks. The `cheese` package was designed with this philosophy in mind, providing tools for working with data and results during statistical analysis.

The dataset used throughout the tutorial comes from the [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Heart+Disease).

```{r}
#Load the package
require(cheese)

#Look at the data
heart_disease

```

Details of the pre-processing steps of the raw data to what is found within the package can be found on the package's Github site [here](https://github.com/zajichek/cheese/blob/master/data-raw/heart_disease.R).

# Summarizing data
```{r}
heart_disease %>%
  univariate_table()

heart_disease %>%
  univariate_table(
    strata = ~ Sex,
    add_n = TRUE
  )
```




