---
title: "Building a descriptive analysis"
output: 
    rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Building a descriptive analysis}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

Once a dataset is cleaned and ready for statistical analysis, the first step is typically to summarize it. The `univariate_table()` function makes it easy to create a custom descriptive analysis while consistently producing clean, presentation-ready output. It is built to integrate directly into your analysis work flow (e.g. R markdown) but can also be called from the console and be rendered in a number of formats.
```{r}
require(cheese)

heart_disease %>%
  univariate_table()

```

By default, an HTML table is produced containing descriptive statistics for columns in the dataset. 

# Custom string templates
In the table above, the summary statistics are presented within the cells in a particular format for different types of data. You can use the `_summary` arguments to customize not only the appearance that the results are presented with, but the values that go into the results themselves.

Suppose instead of the `"median (q1, q3)"` being displayed for numeric data, you want the `"mean [sd] / median"`, in that exact format:
```{r}
heart_disease %>%
  univariate_table(
    numeric_summary = 
      c(
        Summary = "mean [sd] / median"
      )
  )
```

The name `Summary` was used to ensure that the result for the numeric data binded in the same column as the result for the other data types. If you chose to name it something else, you'd get a new column with those summaries:
```{r}
heart_disease %>%
  univariate_table(
    numeric_summary = 
      c(
        NewSummary = "mean [sd] / median"
      )
  )
```

You can add as many summary columns as you want separately for each type of data:
```{r}
heart_disease %>%
  univariate_table(
    numeric_summary = 
      c(
        `Numeric only` = "mean [sd] / median",
        Summary = "median (q1, q3)"
      ),
    categorical_summary = 
      c(
        Summary = "count",
        `Categorical only` = "percent = 100 * proportion"
      )
  )
```

A more visually-appealing case for adding multiple summaries is probably when all the data is the same type:
```{r}
heart_disease %>%
  univariate_table(
    categorical_types = NULL, #Easily disable categorical data from being summarized
    numeric_summary =
      c(
        `Median (Q1, Q3)` = "median (q1, q3)",
        `Min-Max` = "min - max",
        `Mean (SD)` = "mean (sd)"
      )
  )
```

Or when adding a summary that applies to all columns:
```{r}
heart_disease %>%
  univariate_table(
    all_summary = 
      c(
        `# obs. non-missing` = "available of length"
      )
  )
```

These add an extra row for categorical variables. You may have also noticed that the `BloodSugar` column didn't show up in the table until the `all_summary` argument was used--this is because it is not classified as numeric or categorical data, and thus not evaluated by default. See the "Backend functionality" section to learn more.

# Stratification variables
The `strata` argument takes a `formula()` that can be used to stratify the analysis by any number of variables. Columns on the left side will appear down the rows, and columns on the right side will spread across the columns. You can use `+` on either side to specify more than one column. Let's start by stratifying sex across the columns:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ Sex
  )
```

You can do the same thing down the rows:
```{r}
heart_disease %>%
  univariate_table(
    strata = Sex ~ 1
  )
```

Or even both:
```{r}
heart_disease %>%
  univariate_table(
    strata = Sex ~ HeartDisease
  )
```

Now suppose you want both stratification variables across the columns:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ Sex + HeartDisease
  )
```

The levels will span the columns in a hierarchical fashion depending on their order in the formula:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ HeartDisease + Sex
  )
```

Similarly, the rows also collapse hierarchically:
```{r}
heart_disease %>%
  univariate_table(
    strata = HeartDisease + Sex ~ 1
  )
```

You can use any of the functionality described in the previous section with stratification variables as well:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ Sex + HeartDisease,
    numeric_summary = 
      c(
        `Mean (SD)` = "mean (sd)"
      ),
    categorical_summary = 
      c(
        `Count (%)` = "count (percent%)"
      )
  )
```

The summary columns simply get added to the column-spanning hierarchy.

## Adding sample size
The `add_n` argument will add the sample size to the label for the stratification group:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ Sex,
    add_n = TRUE
  )
```

When multiple stratification variables are added on one side of the formula, the sample size will show up on the lowest level of the hierarchy, excluding summary columns:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ Sex + HeartDisease,
    add_n = TRUE
  )
```

A limitation is that when sample size is added in the presence of row and column strata, it is displayed for the marginal groups only:
```{r}
heart_disease %>%
  univariate_table(
    strata = Sex ~ HeartDisease,
    add_n = TRUE
  )
```

## Association metrics
Often when a descriptive analysis is stratified by one or more variables, it is also of interest to add statistics that compare each variable across the groups. The `associations` argument allows you to add a list containing an unlimited number of functions that can produce a scalar value to be placed in the table. First, let's define a function:
```{r}
#Function for a p-value
pval <-
  function(y, x) {
    
    #For categorical data use Fisher's Exact test
    if(some_type(x, "factor")) {
      
      p <- fisher.test(factor(y), factor(x), simulate.p.value = TRUE)$p.value
    
    #Otherwise use Kruskall-Wallis
    } else {
      
      p <- kruskal.test(x, factor(y))$p.value
      
    }
    
    ifelse(p < 0.001, "<0.001", as.character(round(p, 2)))
    
  }

```

The stratification variable will be placed in the _second_ argument of the function(s) provided. Now you can add it to the function call:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ HeartDisease,
    associations = list(`P-value` = pval)
  )
```

The name of function in the list is what becomes the column label. 

The comparison will take place across the number of subgroups there are within the column stratification:
```{r}
heart_disease %>%
  univariate_table(
    strata = ~ Sex + HeartDisease,
    associations = list(`P-value` = pval)
  )
```

However, using a row stratification makes the comparisons be _within_ those groups:
```{r}
heart_disease %>%
  univariate_table(
    strata = Sex ~ HeartDisease,
    associations = list(`P-value` = pval)
  )
```

In general, there must be at least one column stratification variable in order to use association metrics. See `univariate_associations()` for more details on the workhorse of this functionality.

# Backend functionality
`descriptives()` is the function that drives the computation behind the statistics for the columns of the input dataset. Any of its arguments can be passed from `univariate_table()` to add further customization.

## Specifying data types
As noted above, one of columns did not appear in the table by default because it was a `logical()` type. By default, only `factor()` and `numeric()` types are placed into the result, though there are (at least) three ways to include it:

### Change column type prior to function call
You could simply just make the column a conformable type outside of the call:
```{r}
heart_disease %>%
  dplyr::mutate(
    BloodSugar = factor(BloodSugar)
  ) %>%
  univariate_table()
```

### Change scope of what column types are evaluated by what function sets
The `_types` arguments allow you to specify the data types that are to be interpreted by the high-level function call. Let's allow `logical()` types to be treated as a categorical variable:
```{r}
heart_disease %>%
  univariate_table(
    categorical_types = c("factor", "logical")
  )
```

### Allow evaluation by its own set of functions
The most flexible approach would be to define its own set of functions. By default, the data type of anything that is not interpreted as categorical or numeric is considered "other". There is infrastruce in place to supply functions and summaries in the same manner for these columns.
```{r}
heart_disease %>%
  univariate_table(
    f_other = list(count = function(x) table(x)),
    other_summary = 
      c(
        Summary = "count"
      )
  )
```

You would need to also define functions for the percentages, proportions, etc. to exactly match the other examples.

## Adding user-specified functions
You can also add custom functions that can be available for numeric or categorical columns:
```{r}
heart_disease %>%
  univariate_table(
    categorical_types = NULL,
    f_numeric =
      list(
        cv = ~sd(.x) / mean(.x)
      ),
    numeric_summary = 
      c(
        `Coef. of variation` = "sd / mean = cv"
      )
  )
```

The names of functions become the patterns that searched in the string templates.

# Additional preferences
Finally, we'll look at a few of the appearance-related arguments. These can be applied with any combination of other arguments.

## Rendering format
As mentioned above, the default format for the table is HTML, but you could choose an alternative with the `format` argument:
```{r}
heart_disease %>%
  univariate_table(
    format = "none"
  )
```

There are also options for `"latex", "pandoc", "markdown"`.

## Relabeling, releveling and reordering
You can use the `labels` and `levels` arguments to add clean text to any of the variable or categorical level names, and the `order` argument to change the position of the variables in the result:
```{r}
heart_disease %>%
  univariate_table(
    labels = 
      c(
        Age = "Age (years)",
        ChestPain = "Chest pain"
      ),
    levels = 
      list(
        Sex =
          c(
            Male = "M"
          )
      ),
    order = 
      c(
        "BP",
        "Age",
        "Cholesterol"
      )
  )
```

Notice you only need to specify values that need to be changed. Also, ordering is done with the original names even when relabeled.

## Headers, fill values, and captions
The `variableName` and `levelName` arguments are used to change what the headers are for the column names and categorical levels, while `fill_blanks` determines what goes in empty cells. Finally, the `caption` argument specifies labels the entire table:
```{r}
heart_disease %>%
  univariate_table(
    variableName = "THESE ARE VARIABLES",
    levelName = "THESE ARE LEVELS",
    fill_blanks = "BLANK",
    caption = "HERE IS MY CAPTION"
  )
```
